---
title: "Downy mildew assembly quality"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(gt)
library(here)
```

# Data Bases

1. Genomic:
  a. comprehensive database: oomycetes, human, plant, fungi, bacteria (*oomycota-genome*)
  b. contamination database: human, plant, fungi, bacteria (*contam-genome*)
  
2. Protein:
  a. comprehensive database: oomycetes, human, plant, fungi, bacteria (*oomycota-protein*)
  b. contamination database: human, plant, fungi, bacteria (*contam-protein*)

When using comprehensicve database, we extract the sequences that are assign to the oomycete taxa. When using contamination database, we exclude all sequences that mapped to the reference database, and only keep those unmapped sequences.

# Workflow order

1. Classify all the Pacbio sequences with Kraken2 and use metaMDBG to assemble the classified sequences. (*classify-assemble*)

2. Assemble all the Pacbio sequences using metaMDBG, then use Kraken2 to classify the assembled sequences. (*assemble-classify*)

# Confidence score adjustment

definition of confidence score in Kraken2: proportion of matching k-mers assigned to the taxon. The default parameter is 0.

A higher confidence score can reduce misclassification, but also reduce the number of classified sequences. Here, we test four different confidence scores: 0, 0.25, 0.5, 0.75. 



```{r echo=FALSE}
path <- '/Users/yixuanyang/project_data/downy/Kraken-result'
result.dir <- list.dirs(path) %>% str_subset('quality') %>% str_subset('verification', negate = TRUE)
```

BUSCO results are calculated using compleasm pipeline: https://academic.oup.com/bioinformatics/article/39/10/btad595/7284108

```{r echo=FALSE}
read.result <- function(dir) {
    files <- list.files(dir, full.names = TRUE, recursive = TRUE)
    # quast
    info <- str_split(dir, '/') %>% unlist() %>% tail(3)
    info <- data.frame(method = info[1],
                         confidence.score = as.numeric(info[2]))
    if (!any(str_detect(files, 'quast.tsv'))){
      sample.name <- files %>% 
        str_subset('compleasm-eukaryota') %>% 
        basename() %>% 
        str_remove('compleasm-eukaryota-') %>% 
        str_remove('.fasta.gz.txt')
      abb.names <- c('CT', 'LN', 'GC', 'N50')
      quast.sub <- data.frame(sample.name, quast = str_c(abb.names, '0', sep = ':', collapse = '\n'))
    } else {
      quast <- files %>% 
      str_subset('quast.tsv') %>%
        read.csv(sep = '\t') %>%
        as.data.frame() %>% 
        column_to_rownames('Assembly')
        abb.names <- c('CT', 'LN', 'GC', 'N50', 'L50')
      quast.sub <- quast[c(13,15,16,17,20),] %>% 
        apply(2, function(x) str_c(abb.names, x, sep = ':', collapse = '\n')) %>% 
        as.data.frame() %>% 
        rownames_to_column('sample.name')
      names(quast.sub)[2] <- 'quast'
    }
    # compleasm euk
    compleasm.euk <- files %>%
      str_subset('compleasm-eukaryota') %>%
      lapply(function(file) read_lines(file, skip = 1) %>% str_remove(",.*") %>% str_c(collapse = '\n')) %>% 
      unlist()
    sample.name <- files %>% 
      str_subset('compleasm-eukaryota') %>% 
      basename() %>% 
      str_remove('compleasm-eukaryota-') %>% 
      str_remove('.fasta.gz.txt')
    compleasm.euk <- data.frame(sample.name, compleasm.euk) %>% 
      left_join(quast.sub, by = 'sample.name') %>% 
      mutate(quality = str_c(compleasm.euk, quast, sep = '\n')) %>% 
      bind_cols(info)
    # compleasm stram
    compleasm.stramenopiles <- files %>%
      str_subset('compleasm-stramenopiles') %>%
      lapply(function(file) read_lines(file, skip = 1) %>% str_remove(",.*") %>% str_c(collapse = '\n')) %>% 
      unlist()
    sample.name <- files %>%
      str_subset('compleasm-stramenopiles') %>% 
      basename() %>% 
      str_remove('compleasm-stramenopiles-') %>% 
      str_remove('.fasta.gz.txt')
    compleasm.stramenopiles <- data.frame(sample.name, compleasm.stramenopiles) %>% 
      left_join(quast.sub, by = 'sample.name') %>% 
      mutate(quality = str_c(compleasm.stramenopiles, quast, sep = '\n')) %>% 
      bind_cols(info)
    
    result <- list(compleasm.eukaryota = compleasm.euk,
               compleasm.stramenopiles = compleasm.stramenopiles)
    return(result)
}
```

```{r}
euk.result <- map_dfr(result.dir, function(dir) read.result(dir)$compleasm.eukaryota)
stramenopiles.result <- map_dfr(result.dir, function(dir) read.result(dir)$compleasm.stramenopiles)
```

# Assemble quality

## BUSCO: Eukaryota database

```{r echo=FALSE}
tb1 <- euk.result %>% 
    select(method, confidence.score, quality, sample.name) %>%
    group_by(sample.name) %>%
    arrange(sample.name, confidence.score) %>%
    pivot_wider(names_from = method, values_from = quality)
tb1 %>% gt
```

## BUSCO: Stramenopiles database

```{r echo=FALSE}
tb2 <- stramenopiles.result %>% 
    select(method, confidence.score, quality, sample.name) %>%
    group_by(sample.name) %>%
    arrange(sample.name, confidence.score) %>%
    pivot_wider(names_from = method, values_from = quality)
tb2 %>% gt
```



```{r include=FALSE}
fcs <- list.files(path, full.names = TRUE, recursive = TRUE) %>% 
  str_subset('fcs')

fcs.results <- list()
for (name in fcs) {
  fcs.results[[name]] <- read.csv(name, sep = '\t', skip = 1) %>% as.data.frame()
}

fcs.count.tb <- lapply(fcs.results, function(x) x %>% group_by(top_tax_name) %>% summarise(count = n()) %>% arrange(desc(count)))
```

# quality of FCS GX cleaned genomes

```{r echo=FALSE}
path <- '/Users/yixuanyang/project_data/downy/Kraken-result'
result.dir <- list.dirs(path) %>% str_subset('quality') %>% str_subset('verification', negate = FALSE)
new.result.dir <- result.dir %>% str_replace('Kraken-result', 'Kraken-result-cleaned') %>% str_replace('verification/quality', '')
for (i in seq_along(result.dir)){
  if (!dir.exists(new.result.dir[i])) dir.create(new.result.dir[i], recursive = TRUE)
  file.copy(from = result.dir[i], to = new.result.dir[i], overwrite = TRUE, recursive = TRUE)
}

```

```{r}
euk.result <- map_dfr(new.result.dir, function(dir) read.result(dir)$compleasm.eukaryota)
stramenopiles.result <- map_dfr(new.result.dir, function(dir) read.result(dir)$compleasm.stramenopiles)
```


## BUSCO: Eukaryota database

```{r echo=FALSE}
tb1 <- euk.result %>% 
    select(method, confidence.score, quality, sample.name) %>%
    group_by(sample.name) %>%
    arrange(sample.name, confidence.score) %>%
    pivot_wider(names_from = method, values_from = quality)
tb1 %>% gt
```

## BUSCO: Stramenopiles database

```{r echo=FALSE}
tb2 <- stramenopiles.result %>% 
    select(method, confidence.score, quality, sample.name) %>%
    group_by(sample.name) %>%
    arrange(sample.name, confidence.score) %>%
    pivot_wider(names_from = method, values_from = quality)
tb2 %>% gt
```