---
title: "Assembly Benchmark"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
library(gt)
library(here)
library(ggh4x)

source(here("analysis", "lib", "figure-utils.R"))
theme_set(theme_bw())

benchmark_root <- path.expand("~/project_data/downy/benchmarking")
isolate_labels <- c(
  "p_effusa" = "P. Effusa (Reassemble)",
  "MSU1" = "P. Cubensis (MSU1)"
)
qc_table <- here("data", "benchmark_qc", "quality_all_benchmarking.tsv")
if (!file.exists(qc_table)) {
  stop("Could not find quality_all_benchmarking.tsv. Run workflow/03-benchmarking/fasta-quality-table.sh first.")
}

assemblies <- tribble(
  ~isolate, ~method_id, ~method, ~method_dir, ~fasta_rel, ~qc_id,
  "MSU1", "tea_downsample", "target-asm downsample", "tea_downsample", "Quesada_SQIIe_MSU1.fasta.gz", "MSU1_tea_downsample_Quesada_SQIIe_MSU1",
  "MSU1", "tea_no_downsample", "target-asm no downsample", "tea_no_downsample", "Quesada_SQIIe_MSU1.fasta.gz", "MSU1_tea_no_downsample_Quesada_SQIIe_MSU1",
  "MSU1", "hifiasm", "hifiasm", "hifiasm", "MSU1.fasta.gz", "MSU1_hifiasm_MSU1",
  "MSU1", "hifiasm_blastn", "hifiasm + blastn", "hifiasm_blastn", "MSU1.fasta.gz", "MSU1_hifiasm_blastn_MSU1",
  "p_effusa", "tea", "target-asm no downsample", "tea", "p_effusa.fasta.gz", "p_effusa_tea_p_effusa",
  "p_effusa", "hifiasm", "hifiasm", "hifiasm", "p_effusa.fasta.gz", "p_effusa_hifiasm_p_effusa",
  "p_effusa", "hifiasm_blastn", "hifiasm + blastn", "hifiasm_blastn", "p_effusa.fasta.gz", "p_effusa_hifiasm_blastn_p_effusa"
) |>
  mutate(
    fasta_path = file.path(benchmark_root, isolate, method_dir, fasta_rel),
    timing_path = file.path(benchmark_root, isolate, method_dir, "timing.tsv"),
    method = factor(method, levels = c("target-asm no downsample", "target-asm downsample", "hifiasm", "hifiasm + blastn"))
  )

missing_assemblies <- assemblies |>
  filter(!file.exists(fasta_path))
```

This report compares the benchmarking assemblies in `r benchmark_root`. Completeness and assembly statistics are read from `r qc_table`, generated by `workflow/03-benchmarking/fasta-quality-table.sh`.

```{r missing-assemblies-note, echo=FALSE, results='asis'}
if (nrow(missing_assemblies)) {
  cat("> Missing assembly FASTA files. Re-run the corresponding benchmark method before using those rows: ",
      paste(paste0(missing_assemblies$isolate, ": ", missing_assemblies$method, " (", missing_assemblies$fasta_path, ")"), collapse = "; "),
      ".\n", sep = "")
}
```

```{r read-data, echo=FALSE, message=FALSE, warning=FALSE}
read_timing <- function(path) {
  if (!file.exists(path) || file.info(path)$size == 0) {
    return(tibble(total_seconds = NA_real_))
  }

  tb <- suppressMessages(read_tsv(path, show_col_types = FALSE))
  total <- tb |>
    filter(step == "Total") |>
    slice_tail(n = 1) |>
    pull(seconds)

  total <- if (length(total)) total else na_if(sum(tb$seconds, na.rm = TRUE), 0)

  tibble(total_seconds = as.numeric(total))
}

assembly_stats <- assemblies |>
  bind_cols(map_dfr(assemblies$timing_path, read_timing)) |>
  mutate(
    runtime_hours = total_seconds / 3600,
    runtime_label = if_else(
      is.na(runtime_hours),
      "not recorded",
      sprintf("%.2f h", runtime_hours)
    )
  )

quality <- suppressMessages(read_tsv(qc_table, show_col_types = FALSE)) |>
  rename(qc_id = file) |>
  mutate(
    single = parse_number(S),
    duplicated = parse_number(D),
    fragmented = parse_number(F),
    missing = parse_number(M),
    contigs = as.integer(.data[["# contigs"]]),
    total_length = as.numeric(.data[["Total length"]]),
    n50 = as.numeric(N50)
  )

benchmark <- assembly_stats |>
  left_join(quality, by = "qc_id") |>
  mutate(
    complete = single + duplicated,
    isolate_label = recode(isolate, !!!isolate_labels)
  )

method_benchmark <- benchmark |>
  filter(method_id %in% c("hifiasm", "hifiasm_blastn", "tea", "tea_no_downsample")) |>
  mutate(
    method_compare = case_when(
      method_id %in% c("tea", "tea_no_downsample") ~ "target-asm",
      TRUE ~ as.character(method)
    ),
    method_compare = factor(method_compare, levels = c("hifiasm", "hifiasm + blastn", "target-asm"))
  )

target_asm_sensitivity <- benchmark |>
  filter(isolate == "MSU1", method_id %in% c("tea_downsample", "tea_no_downsample")) |>
  mutate(
    method_compare = factor(
      method_id,
      levels = c("tea_downsample", "tea_no_downsample"),
      labels = c("Downsampled", "All reads")
    )
  )

make_metric_matrix_data <- function(data) {
  completeness_data <- data |>
    transmute(
      isolate,
      isolate_label,
      method_compare,
      single,
      duplicated,
      complete,
      metric = "Completeness",
      annotation = sprintf("S: %.1f%%\nD: %.1f%%", single, duplicated)
    ) |>
    pivot_longer(c(single, duplicated), names_to = "component", values_to = "value") |>
    mutate(
      component = recode(component, single = "Single", duplicated = "Duplicated"),
      fill_group = component
    )

  metric_data <- data |>
    transmute(
      isolate,
      isolate_label,
      method_compare,
      `Number of contigs` = contigs,
      Runtime = runtime_hours
    ) |>
    pivot_longer(c(`Number of contigs`, Runtime), names_to = "metric", values_to = "value") |>
    mutate(
      component = "Value",
      fill_group = "Value",
      complete = NA_real_,
      annotation = NA_character_
    )

  bind_rows(completeness_data, metric_data) |>
    mutate(
      metric = factor(metric, levels = c("Completeness", "Number of contigs", "Runtime")),
      isolate_label = factor(isolate_label, levels = c("P. Effusa (Reassemble)", "P. Cubensis (MSU1)"))
    )
}

make_metric_matrix_plot <- function(data, rows = vars(isolate_label), x_text_angle = 25) {
  plot_data <- make_metric_matrix_data(data)
  labels <- plot_data |>
    filter(metric == "Completeness") |>
    distinct(isolate_label, method_compare, metric, annotation) |>
    mutate(y = 101.5)

  ggplot(plot_data, aes(x = method_compare, y = value, fill = fill_group)) +
    geom_col(width = 0.68, na.rm = TRUE) +
    geom_text(
      data = labels,
      aes(x = method_compare, y = y, label = annotation),
      inherit.aes = FALSE,
      size = 2.8,
      lineheight = 0.9
    ) +
    facet_grid2(
      rows = rows,
      cols = vars(metric),
      scales = "free_y",
      independent = "y",
      axes = "y"
    ) +
    facetted_pos_scales(
      y = list(
        metric == "Completeness" ~ scale_y_continuous(
          limits = c(50, 104),
          breaks = seq(50, 100, 10),
          labels = scales::label_percent(scale = 1),
          oob = scales::oob_keep
        ),
        metric == "Number of contigs" ~ scale_y_continuous(
          labels = scales::label_number(big.mark = ","),
          expand = expansion(mult = c(0, 0.08))
        ),
        metric == "Runtime" ~ scale_y_continuous(
          labels = scales::label_number(suffix = " h"),
          expand = expansion(mult = c(0, 0.08))
        )
      )
    ) +
    labs(x = NULL, y = NULL) +
    scale_fill_manual(values = c(
      "Single" = "#0072B2",
      "Duplicated" = "#E69F00",
      "Value" = "#56B4E9"
    )) +
    guides(fill = "none") +
    theme(
      axis.text.x = element_text(angle = x_text_angle, hjust = if (x_text_angle == 0) 0.5 else 1),
      legend.position = "none"
    )
}
```

## Assembly Method Comparison

This section compares `hifiasm`, `hifiasm + blastn`, and target-asm. For P. Cubensis (MSU1), the target-asm row uses the no-downsampling run.

```{r method-summary-table, echo=FALSE}
method_benchmark |>
  select(isolate = isolate_label, method = method_compare, single, duplicated, complete, contigs, total_length, n50, runtime_hours) |>
  arrange(isolate, method) |>
  gt(groupname_col = "isolate") |>
  fmt_number(columns = c(single, duplicated, complete), decimals = 2) |>
  fmt_number(columns = c(contigs, total_length, n50), decimals = 0) |>
  fmt_number(columns = runtime_hours, decimals = 2) |>
  cols_label(
    method = "Method",
    single = "Single BUSCO (%)",
    duplicated = "Duplicated BUSCO (%)",
    complete = "Complete BUSCO (%)",
    contigs = "Contigs",
    total_length = "Total length",
    n50 = "N50",
    runtime_hours = "Runtime (h)"
  )
```

```{r method-matrix-plot, echo=FALSE, fig.width=7.2, fig.height=4.7}
method_matrix_plot <- make_metric_matrix_plot(method_benchmark)
save_pub_r(method_matrix_plot, "benchmark-method-matrix", width_mm = 180, height_mm = 120)
method_matrix_plot
```

```{r missing-runtime-note, echo=FALSE, results='asis'}
missing_runtime <- method_benchmark |>
  filter(is.na(runtime_hours)) |>
  mutate(label = paste(isolate, method_compare, sep = ": ")) |>
  pull(label)

if (length(missing_runtime)) {
  cat("Runtime was not recorded for: ", paste(missing_runtime, collapse = "; "), ".\n", sep = "")
}
```

## target-asm Downsampling Sensitivity

This section compares the two P. Cubensis (MSU1) target-asm runs: one using the downsampled reads and one using all mapped reads.

```{r tea-sensitivity-summary, echo=FALSE}
target_asm_sensitivity |>
  select(method = method_compare, single, duplicated, complete, contigs, total_length, n50, runtime_hours) |>
  arrange(method) |>
  gt() |>
  fmt_number(columns = c(single, duplicated, complete), decimals = 2) |>
  fmt_number(columns = c(contigs, total_length, n50), decimals = 0) |>
  fmt_number(columns = runtime_hours, decimals = 2) |>
  cols_label(
    method = "target-asm run",
    single = "Single BUSCO (%)",
    duplicated = "Duplicated BUSCO (%)",
    complete = "Complete BUSCO (%)",
    contigs = "Contigs",
    total_length = "Total length",
    n50 = "N50",
    runtime_hours = "Runtime (h)"
  )
```

```{r tea-sensitivity-plot, echo=FALSE, fig.width=7.2, fig.height=3.0}
target_asm_sensitivity_plot <- make_metric_matrix_plot(target_asm_sensitivity, rows = vars(), x_text_angle = 0)
save_pub_r(target_asm_sensitivity_plot, "benchmark-target-asm-sensitivity", width_mm = 180, height_mm = 75)
target_asm_sensitivity_plot
```
