---
title: "Assembly Benchmark"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(gt)
library(here)
library(ggh4x)

theme_set(theme_bw())

benchmark_root <- file.path(path.expand(Sys.getenv("PROJECT_DATA", "~/project_data/downy")), "results/benchmarking")
qc_table <- here("data", "benchmark_qc", "quality_all_benchmarking.tsv")
isolate_labels <- c("UA202013" = "P. effusa (UA202013)", "MSU1" = "P. cubensis (MSU1)")

# qc_id is the FASTA identifier written by workflow/benchmarking/fasta-quality-table.sh.
assemblies <- tribble(
  ~isolate, ~method_id, ~qc_id,
  "MSU1", "targetasm_downsample", "MSU1_targetasm_downsample_Quesada_SQIIe_MSU1",
  "MSU1", "targetasm_no_downsample", "MSU1_targetasm_no_downsample_Quesada_SQIIe_MSU1",
  "MSU1", "hifiasm", "MSU1_hifiasm_MSU1",
  "MSU1", "hifiasm_blastn", "MSU1_hifiasm_blastn_MSU1",
  "UA202013", "targetasm_no_downsample", "UA202013_targetasm_no_downsample_UA202013",
  "UA202013", "hifiasm", "UA202013_hifiasm_UA202013",
  "UA202013", "hifiasm_blastn", "UA202013_hifiasm_blastn_UA202013"
) |>
  mutate(timing_path = file.path(benchmark_root, isolate, method_id, "timing.tsv"))
```

This report compares the benchmarking assemblies in `r benchmark_root`. Completeness and assembly statistics are read from `r qc_table`, generated by `workflow/benchmarking/fasta-quality-table.sh`.

```{r read-data, echo=FALSE, message=FALSE, warning=FALSE}
read_total_hours <- function(path) {
  total <- read_tsv(path, show_col_types = FALSE) |>
    filter(step == "Total") |>
    pull(seconds)
  if (length(total) != 1 || !is.finite(total) || total <= 0) stop("Invalid Total runtime: ", path)
  total / 3600
}

quality <- read_tsv(qc_table, show_col_types = FALSE) |>
  transmute(
    qc_id = file,
    single = parse_number(S),
    duplicated = parse_number(D),
    complete = single + duplicated,
    contigs = as.integer(`# contigs`),
    total_length = as.numeric(`Total length`),
    n50 = as.numeric(N50)
  )

benchmark <- assemblies |>
  mutate(
    runtime_hours = map_dbl(timing_path, read_total_hours),
    isolate_label = recode(isolate, !!!isolate_labels)
  ) |>
  left_join(quality, by = "qc_id")

stopifnot(nrow(quality) == nrow(assemblies), !anyNA(benchmark$contigs), !anyNA(benchmark$single))

# Method comparison: targetasm without downsampling against the two hifiasm baselines.
method_benchmark <- benchmark |>
  filter(method_id != "targetasm_downsample") |>
  mutate(method_compare = factor(
    method_id,
    levels = c("hifiasm", "hifiasm_blastn", "targetasm_no_downsample"),
    labels = c("hifiasm", "hifiasm + blastn", "targetasm")
  ))

# Downsampling sensitivity: the two MSU1 targetasm runs.
targetasm_sensitivity <- benchmark |>
  filter(isolate == "MSU1", str_starts(method_id, "targetasm")) |>
  mutate(method_compare = factor(
    method_id,
    levels = c("targetasm_downsample", "targetasm_no_downsample"),
    labels = c("Downsampled", "All reads")
  ))

make_metric_matrix_data <- function(data) {
  completeness_data <- data |>
    transmute(
      isolate,
      isolate_label,
      method_compare,
      single,
      duplicated,
      complete,
      metric = "Completeness",
      annotation = sprintf("S: %.1f%%\nD: %.1f%%", single, duplicated)
    ) |>
    pivot_longer(c(single, duplicated), names_to = "component", values_to = "value") |>
    mutate(
      component = recode(component, single = "Single", duplicated = "Duplicated"),
      fill_group = component
    )

  metric_data <- data |>
    transmute(
      isolate,
      isolate_label,
      method_compare,
      `Number of contigs` = contigs,
      Runtime = runtime_hours
    ) |>
    pivot_longer(c(`Number of contigs`, Runtime), names_to = "metric", values_to = "value") |>
    mutate(
      component = "Value",
      fill_group = "Value",
      complete = NA_real_,
      annotation = NA_character_
    )

  bind_rows(completeness_data, metric_data) |>
    mutate(
      metric = factor(metric, levels = c("Completeness", "Number of contigs", "Runtime")),
      isolate_label = factor(isolate_label, levels = unname(isolate_labels))
    )
}

make_metric_matrix_plot <- function(data, rows = vars(isolate_label), x_text_angle = 25) {
  plot_data <- make_metric_matrix_data(data)
  labels <- plot_data |>
    filter(metric == "Completeness") |>
    distinct(isolate_label, method_compare, metric, annotation) |>
    mutate(y = 101.5)

  ggplot(plot_data, aes(x = method_compare, y = value, fill = fill_group)) +
    geom_col(width = 0.68, na.rm = TRUE) +
    geom_text(
      data = labels,
      aes(x = method_compare, y = y, label = annotation),
      inherit.aes = FALSE,
      size = 2.8,
      lineheight = 0.9
    ) +
    facet_grid2(
      rows = rows,
      cols = vars(metric),
      scales = "free_y",
      independent = "y",
      axes = "y"
    ) +
    facetted_pos_scales(
      y = list(
        metric == "Completeness" ~ scale_y_continuous(
          limits = c(50, 104),
          breaks = seq(50, 100, 10),
          labels = scales::label_percent(scale = 1),
          oob = scales::oob_keep
        ),
        metric == "Number of contigs" ~ scale_y_continuous(
          labels = scales::label_number(big.mark = ","),
          expand = expansion(mult = c(0, 0.08))
        ),
        metric == "Runtime" ~ scale_y_continuous(
          labels = scales::label_number(suffix = " h"),
          expand = expansion(mult = c(0, 0.08))
        )
      )
    ) +
    labs(x = NULL, y = NULL) +
    scale_fill_manual(values = c(
      "Single" = "#0072B2",
      "Duplicated" = "#E69F00",
      "Value" = "#56B4E9"
    )) +
    guides(fill = "none") +
    theme(
      axis.text.x = element_text(angle = x_text_angle, hjust = if (x_text_angle == 0) 0.5 else 1),
      legend.position = "none"
    )
}
```

## Assembly Method Comparison

This section compares `hifiasm`, `hifiasm + blastn`, and targetasm. For P. cubensis (MSU1), the targetasm row uses the no-downsampling run.

```{r method-summary-table, echo=FALSE}
method_benchmark |>
  select(isolate = isolate_label, method = method_compare, single, duplicated, complete, contigs, total_length, n50, runtime_hours) |>
  arrange(isolate, method) |>
  gt(groupname_col = "isolate") |>
  fmt_number(columns = c(single, duplicated, complete), decimals = 2) |>
  fmt_number(columns = c(contigs, total_length, n50), decimals = 0) |>
  fmt_number(columns = runtime_hours, decimals = 2) |>
  cols_label(
    method = "Method",
    single = "Single BUSCO (%)",
    duplicated = "Duplicated BUSCO (%)",
    complete = "Complete BUSCO (%)",
    contigs = "Contigs",
    total_length = "Total length",
    n50 = "N50",
    runtime_hours = "Runtime (h)"
  )
```

```{r method-matrix-plot, echo=FALSE, fig.width=7.2, fig.height=4.7}
method_matrix_plot <- make_metric_matrix_plot(method_benchmark)
out <- here("figures", "benchmark-method-matrix")
ggsave(paste0(out, ".svg"), method_matrix_plot, device = svglite::svglite, width = 180, height = 120, units = "mm")
ggsave(paste0(out, ".pdf"), method_matrix_plot, device = grDevices::pdf, width = 180, height = 120, units = "mm", family = "Helvetica", useDingbats = FALSE)
ggsave(paste0(out, ".tiff"), method_matrix_plot, device = ragg::agg_tiff, width = 180, height = 120, units = "mm", dpi = 600, compression = "lzw")
ggsave(paste0(out, ".png"), method_matrix_plot, device = ragg::agg_png, width = 180, height = 120, units = "mm", dpi = 600)
method_matrix_plot
```

## targetasm Downsampling Sensitivity

This section compares the two P. cubensis (MSU1) targetasm runs: one using the downsampled reads and one using all mapped reads.

```{r targetasm-sensitivity-summary, echo=FALSE}
targetasm_sensitivity |>
  select(method = method_compare, single, duplicated, complete, contigs, total_length, n50, runtime_hours) |>
  arrange(method) |>
  gt() |>
  fmt_number(columns = c(single, duplicated, complete), decimals = 2) |>
  fmt_number(columns = c(contigs, total_length, n50), decimals = 0) |>
  fmt_number(columns = runtime_hours, decimals = 2) |>
  cols_label(
    method = "targetasm run",
    single = "Single BUSCO (%)",
    duplicated = "Duplicated BUSCO (%)",
    complete = "Complete BUSCO (%)",
    contigs = "Contigs",
    total_length = "Total length",
    n50 = "N50",
    runtime_hours = "Runtime (h)"
  )
```

```{r targetasm-sensitivity-plot, echo=FALSE, fig.width=7.2, fig.height=3.0}
targetasm_sensitivity_plot <- make_metric_matrix_plot(targetasm_sensitivity, rows = vars(), x_text_angle = 0)
out <- here("figures", "benchmark-targetasm-sensitivity")
ggsave(paste0(out, ".svg"), targetasm_sensitivity_plot, device = svglite::svglite, width = 180, height = 75, units = "mm")
ggsave(paste0(out, ".pdf"), targetasm_sensitivity_plot, device = grDevices::pdf, width = 180, height = 75, units = "mm", family = "Helvetica", useDingbats = FALSE)
ggsave(paste0(out, ".tiff"), targetasm_sensitivity_plot, device = ragg::agg_tiff, width = 180, height = 75, units = "mm", dpi = 600, compression = "lzw")
ggsave(paste0(out, ".png"), targetasm_sensitivity_plot, device = ragg::agg_png, width = 180, height = 75, units = "mm", dpi = 600)
targetasm_sensitivity_plot
```
