---
title: "Final Assembly"
author: "Yixuan Yang"
date: "`r Sys.Date()`"
output: html_document
---

```{r echo=FALSE, message=FALSE}
library(GENESPACE)
library(here)
library(tidyverse)
library(Biostrings)
library(data.table)
library(pctax)
library(pcutils)
library(gt)
theme_set(theme_bw())
```

```{r echo=FALSE, message=FALSE}
read.tb <- function(path, method) {
  files <- list.files(path, full.names = TRUE, pattern = 'quality.csv', recursive = TRUE)
  tb <- map_dfr(files, function(x) {data.frame(fread(x))})
  return(tb)
}
```

After synteny with P.effusa, we have new assembly files where some of the contigs are glued. Let's check their qualities.

```{r echo=FALSE}
read.tb('~/project_data/downy/cleaned_contigs', 'final assembly') |> 
  cbind(sample = c('p_effusa', 'Quessda_MSU1','Quessda_Phumuli','Quessda_SC1982'))|>
  relocate(sample) |> 
  gt() |> fmt_number()
```

# Read Length

```{r echo=FALSE}
read.length <- list.files('~/project_data/downy/cleaned_contigs', pattern = '.fasta.gz', full.names = TRUE, recursive = TRUE) |> 
  map_dfr(function(file) {
    seq <- readDNAStringSet(file)
    data.frame(sample = str_remove(basename(file), '.fasta.gz'),
               contig = names(seq),
               length = width(seq))
    }) |> group_by(sample) |> 
  arrange(desc(length)) |>
  top_n(30) |> 
  mutate(rank = 1:n()) |> 
  mutate(ragtag = if_else(str_detect(contig, '_RagTag'), 'scaffold' ,'unplaced contigs')) |> 
  ungroup()

read.length$sample <- factor(read.length$sample, levels = c("p_effusa", "Quesada_SQIIe_SC1982", "Quesada_SQIIe_MSU1", "Quesada_SQIIe_Phumuli"))

ggplot(read.length, aes(x = factor(rank), y = length, fill = ragtag)) +
  geom_col() +
  scale_y_continuous(labels = scales::label_number(big.mark = ",")) +
  facet_wrap(~sample, scales = 'free_y', ncol = 1) +
  labs(
    title = "Top 30 Longest Contigs",
    subtitle = "Ordered by length",
    x = "Rank",
    y = "Length (bp)"
  )
```


# Syntenic Plots

Here, we can run orthofinder and genespace to check the synteny of each isolate.

The orthofinder we used is:

```{bash eval=FALSE}
orthofinder -f "$HOME/project_data/downy/genespace/tmp" -t 10 -a 10 -X -o "$HOME/project_data/downy/genespace/orthofinder"
```

Reference genomes: Peronospora effusa, Bremia lactucae, Phytophthora sojae

Our isolates: Quesada_SQIIe_MSU1, Quesada_SQIIe_SC1982, Quesada_SQIIe_Phumuli

```{r  message=FALSE, echo=FALSE, results=FALSE}
genomeRepo <- "~/project_data/downy/genespace/genomes/"
wd <- "~/project_data/downy/genespace/"
path2mcscanx <- "/Users/yixuanyang/miniforge3/bin/"

genomes2run <- c("Peronospora_effusa", "Bremia_lactucae", "Phytophthora_sojae")
parsedPaths <- parse_annotations(
  rawGenomeRepo = genomeRepo,
  genomeDirs = genomes2run,
  genomeIDs = genomes2run,
  headerSep = NA,
  headerEntryIndex = 1,
  genespaceWd = wd)
asm2run <- c("Quesada_SQIIe_SC1982", "Quesada_SQIIe_MSU1", "Quesada_SQIIe_Phumuli")
parsedPaths <- parse_annotations(
  rawGenomeRepo = genomeRepo,
  genomeDirs = asm2run,
  genomeIDs = asm2run,
  headerSep = NA,
  headerEntryIndex = 1,
  gffIdColumn = "ID",
  genespaceWd = wd)

gpar <- init_genespace(
  wd = wd,
  nCores = 10,
  dotplots = 'never',
  path2mcscanx = path2mcscanx)

out <- run_genespace(gpar)
```

# Riparian

```{r echo=FALSE}
for (asm in asm2run){
  plot_riparian(
    gsParam = out,
    genomeIDs = c(asm, genomes2run[1]),
    refGenome = genomes2run[1],
    chrLabFontSize = 2,
    useRegions = FALSE)
}
```

```{r echo=FALSE}
full.plot <- plot_riparian(
  gsParam = out,
  genomeIDs = c(genomes2run[1],asm2run),
  refGenome = genomes2run[1],
  chrLabFontSize = 2,
  useRegions = FALSE)
```

```{r echo=FALSE}
full.plot <- plot_riparian(
  gsParam = out,
  genomeIDs = c(genomes2run,asm2run),
  refGenome = genomes2run[1],
  chrLabFontSize = 2,
  useRegions = FALSE)
```


```{r echo=FALSE, include=FALSE}
for (ref in genomes2run){
  for (asm in asm2run){
    plot_riparian(
      gsParam = out,
      genomeIDs = c(asm, ref),
      refGenome = asm,
      pdfFile = here('figures', str_c(asm, '-vs-', ref, '.pdf')),
      chrLabFontSize = 3,
      useRegions = FALSE)
  }
}

```

```{r echo=FALSE, include=FALSE}
syntenicRegion <- file.path(wd, 'results','syntenicBlock_coordinates.csv') |> 
  fread() |> filter((genome1 %in% asm2run)& genome2 == "Peronospora_effusa") |>
  group_by(genome1)

syntenicRegion |>
  summarise(unique_chr_count = n_distinct(chr1)) |> 
  gt()
```

# Phylogenetic Tree

```{r echo=FALSE, message=FALSE}
library(ggtree)
tree.path <- file.path(wd, 'results/SpeciesTree_rooted.txt')
ggtree(read.tree(tree.path)) + geom_tiplab(align = T)  + xlim(0, 0.6)
```




