---
title: "Final Assembly"
author: "Yixuan Yang"
date: "`r Sys.Date()`"
output: html_document
---

```{r echo=FALSE, message=FALSE}
library(GENESPACE)
library(here)
library(tidyverse)
library(Biostrings)
library(data.table)
library(gt)
theme_set(theme_bw())
```

```{r echo=FALSE, message=FALSE}
read.tb <- function(path, method) {
  files <- list.files(path, full.names = TRUE, pattern = 'stram.csv')
  tb <- map_dfr(files, function(x) {data.frame(method = method,
                                               sample = str_remove(basename(x), '_stram.csv'),
                                               fread(x))})
  return(tb)
}
```

After synteny with P.effusa, we have new assembly files where some of the contigs are glued. Let's check their qualities.

```{r echo=FALSE}
read.tb('~/project_data/downy/hifiasm/fcs-gx/kraken2/purge_dups/rag_tag/compleasm/', 'final assembly') |> gt() |> fmt_number()
```

# Read Length

```{r echo=FALSE}
read.length <- list.files('~/project_data/downy/hifiasm/fcs-gx/kraken2/purge_dups/rag_tag', pattern = '.fasta.gz', full.names = TRUE) |> 
  map_dfr(function(file) {
    seq <- readDNAStringSet(file)
    data.frame(sample = str_remove(basename(file), '.fasta.gz'),
               contig = names(seq),
               length = width(seq))
    }) |> group_by(sample) |> 
  arrange(desc(length)) |>
  top_n(30) |> 
  mutate(rank = 1:n()) |> 
  mutate(is_scaffolded = str_detect(contig, '_RagTag')) |> 
  ungroup()

ggplot(read.length, aes(x = factor(rank), y = length, fill = is_scaffolded)) +
  geom_col() +
  scale_y_continuous(labels = scales::label_number(big.mark = ",")) +
  facet_wrap(~sample, scales = 'free_y', ncol = 1) +
  labs(
    title = "Top 30 Longest Contigs",
    subtitle = "Ordered by length",
    x = "Rank",
    y = "Length (bp)"
  )
```

We see there's one long contig in SC1982 that is not glued (or there's no cds sequence detected by helixer). `ptg000299l_1` is the name of that contig.

```{r echo=FALSE}
(read.length |> filter(sample == 'Quesada_SQIIe_SC1982'))[17,]
```


# Syntenic Plots

Here, we can run orthofinder and genespace to check the synteny of each isolate.

The orthofinder we used is:

```{bash eval=FALSE}
orthofinder -f /Users/yixuanyang/project_data/downy/genespace/tmp -t 10 -a 1 -X -o /Users/yixuanyang/project_data/downy/genespace/orthofinder
```

Reference genomes: Peronospora effusa, Bremia lactucae, Phytophthora sojae

Our isolates: Quesada_SQIIe_MSU1, Quesada_SQIIe_SC1982, Quesada_SQIIe_Phumuli

```{r  message=FALSE, echo=FALSE, results=FALSE}
genomeRepo <- "~/project_data/downy/genespace/genomes/"
wd <- "~/project_data/downy/genespace/"
path2mcscanx <- "/Users/yixuanyang/miniforge3/bin/"

genomes2run <- c("Peronospora_effusa", "Bremia_lactucae", "Phytophthora_sojae")
parsedPaths <- parse_annotations(
  rawGenomeRepo = genomeRepo,
  genomeDirs = genomes2run,
  genomeIDs = genomes2run,
  headerSep = NA,
  headerEntryIndex = 1,
  genespaceWd = wd)
asm2run <- c("Quesada_SQIIe_MSU1", "Quesada_SQIIe_Phumuli", "Quesada_SQIIe_SC1982")
parsedPaths <- parse_annotations(
  rawGenomeRepo = genomeRepo,
  genomeDirs = asm2run,
  genomeIDs = asm2run,
  headerSep = NA,
  headerEntryIndex = 1,
  gffIdColumn = "ID",
  genespaceWd = wd)

gpar <- init_genespace(
  wd = wd,
  nCores = 10,
  dotplots = 'never',
  path2mcscanx = path2mcscanx)

out <- run_genespace(gpar)
```

# Riparian

```{r echo=FALSE}
for (asm in asm2run){
  plot_riparian(
    gsParam = out,
    genomeIDs = c(asm, genomes2run),
    refGenome = "Peronospora_effusa",
    chrLabFontSize = 2,
    useRegions = FALSE)
}
```

Number of contigs mapped to P.effusa

```{r echo=FALSE}
syntenicRegion <- file.path(wd, 'results','syntenicBlock_coordinates.csv') |> 
  fread() |> filter((genome1 %in% asm2run)& genome2 == "Peronospora_effusa") |>
  group_by(genome1)

syntenicRegion |>
  summarise(unique_chr_count = n_distinct(chr1)) |> 
  gt()
```


