---
title: "Final Assembly"
author: "Yixuan Yang"
date: "`r Sys.Date()`"
output: html_document
---

```{r echo=FALSE, message=FALSE}
library(here)
library(tidyverse)
library(Biostrings)
library(data.table)
library(gt)
knitr::opts_knit$set(root.dir = here())
theme_set(theme_bw())
```

```{r echo=FALSE, message=FALSE}
read.tb <- function(path, method) {
  files <- list.files(path, full.names = TRUE, pattern = 'quality_final.csv', recursive = TRUE)
  tb <- map_dfr(files, function(x) {data.frame(fread(x))})
  return(tb)
}
```

```{r echo=FALSE}
read.tb('~/project_data/downy/Assembly/', 'final assembly') |> 
  cbind(sample = c('p_effusa', 'Quessda_MSU1','Quessda_Phumuli','Quessda_SC1982'))|>
  relocate(sample) |> 
  gt() |> fmt_number()
```


# Read Length

```{r echo=FALSE}
read.length <- list.files('~/project_data/downy/Assembly/',
                          pattern = '.fasta.gz', 
                          full.names = TRUE, recursive = TRUE) |> 
  (\(x) x[!str_detect(x, "fcs_gx|hifiasm|metaMDBG")])() |> 
  map_dfr(function(file) {
    seq <- readDNAStringSet(file)
    data.frame(sample = str_remove(basename(file), '.fasta.gz'),
               contig = names(seq),
               length = width(seq))
    }) |> group_by(sample) |> 
  arrange(desc(length)) |>
  top_n(30) |> 
  mutate(rank = 1:n()) |> 
  ungroup()

read.length$sample <- factor(read.length$sample, levels = c("p_effusa", "Quesada_SQIIe_SC1982", "Quesada_SQIIe_MSU1", "Quesada_SQIIe_Phumuli"))

ggplot(read.length, aes(x = factor(rank), y = length)) +
  geom_col() +
  scale_y_continuous(labels = scales::label_number(big.mark = ",")) +
  facet_wrap(~sample, scales = 'free_y', ncol = 1) +
  labs(
    title = "Top 30 Longest Contigs",
    subtitle = "Ordered by length",
    x = "Rank",
    y = "Length (bp)"
  )
```


# Rename the contigs

We are going to rename the contigs in a consistent naming logic: species_sizeRank.

The corresponding names are shown below:

| Sample | Sequence name |
|---|---|
| p_effusa | Peff-26 |
| Quesada_SQIIe_MSU1 | Pcub-MSU1 |
| Quesada_SQIIe_SC1982 | Pcub-SC1982 |
| Quesada_SQIIe_Phumuli | Phum-OR502AA |

```{r echo=FALSE, message=FALSE}
rename_key <- tribble(
  ~old_file, ~seq_name, ~file_name,
  "p_effusa", "Peff-26", "Peronospora_effusa_reassemble",
  "Quesada_SQIIe_MSU1", "Pcub-MSU1", "Pseudoperonospora_cubensis_MSU1",
  "Quesada_SQIIe_SC1982", "Pcub-SC1982", "Pseudoperonospora_cubensis_SC1982",
  "Quesada_SQIIe_Phumuli", "Phum-OR502AA", "Pseudoperonospora_humuli_OR502AA"
)

rename_dir <- path.expand("~/project_data/downy/contigs-renamed")
dir.create(rename_dir, recursive = TRUE, showWarnings = FALSE)

assembly_files <- list.files(
  "~/project_data/downy/Assembly/",
  pattern = "\\.fasta\\.gz$",
  full.names = TRUE,
  recursive = TRUE
) |>
  (\(x) x[!str_detect(x, "fcs_gx|hifiasm|metaMDBG")])()

rename_contigs_by_length <- function(file) {
  input_name <- str_remove(basename(file), "\\.fasta\\.gz$")
  row_id <- match(input_name, rename_key$old_file)
  stopifnot(!is.na(row_id))

  seqs <- readDNAStringSet(file)
  seqs <- seqs[order(width(seqs), decreasing = TRUE)]
  names(seqs) <- paste0(rename_key$seq_name[row_id], "_", str_pad(seq_along(seqs), width = 3, pad = "0"))

  writeXStringSet(
    seqs,
    file.path(rename_dir, paste0(rename_key$file_name[row_id], ".fasta.gz")),
    compress = TRUE
  )
}

walk(assembly_files, rename_contigs_by_length)
```


# Remove mitochondrial contigs

After assembly, some mitochondrial contigs were still retained in the assembly files. After renaming the contigs, we identified these mitochondrial contigs by aligning each assembly against the mitochondrial genome, then separated them from the rest of the assembly.

```{bash echo=TRUE}
set -euo pipefail

INPUT_DIR="${HOME}/project_data/downy/contigs-renamed"
RESULT_DIR="${HOME}/project_data/downy/contigs-renamed/mitochondiral"
MITO_REF="${HOME}/project_data/downy/mitochondrial-genome/KT072718.1.fna"

mkdir -p "$RESULT_DIR"

find "$INPUT_DIR" -maxdepth 1 -type f -name "*.fasta.gz" | while read -r FILE; do
  BASENAME="$(basename "$FILE" .fasta.gz)"
  OUT="$RESULT_DIR/${BASENAME}.mito.tsv"

  if [ -s "$OUT" ]; then
    echo "Skipping existing: $OUT"
    continue
  fi

  TMP="$RESULT_DIR/${BASENAME}.fasta"
  gzip -dc "$FILE" > "$TMP"

  blastn \
    -query "$TMP" \
    -subject "$MITO_REF" \
    -outfmt "6 qseqid sseqid pident length qlen slen evalue" \
    -max_target_seqs 1 \
    -max_hsps 1 \
    -out "$OUT"

  rm "$TMP"
done
```

```{r echo=FALSE, message=FALSE}
blast_files <- list.files(
  "~/project_data/downy/contigs-renamed/mitochondiral",
  pattern = "\\.mito\\.tsv$",
  full.names = TRUE
)

read_mito_blast <- function(file) {
  if (file.info(file)$size == 0) {
    return(tibble())
  }

  fread(
    file,
    col.names = c("seqid", "subject", "pident", "align.length",
                  "query.length", "subject.length", "e.value")
  ) |>
    as_tibble() |>
    mutate(sample = str_remove(basename(file), "\\.mito\\.tsv$"))
}

blast.mito <- map_dfr(blast_files, read_mito_blast) |>
  relocate(sample) |>
  mutate(`align/query` = round(align.length / query.length, 3))

blast.mito |>
  group_by(sample) |>
  arrange(desc(`align/query`), desc(align.length), .by_group = TRUE) |>
  gt()
```


We treated contigs as mitochondrial when at least 10% of the contig aligned to the mitochondrial genome.

```{r echo=FALSE, message=FALSE}
mito_contigs <- blast.mito |>
  filter(`align/query` >= 0.1)

mito_contigs |>
  group_by(sample) |>
  arrange(desc(`align/query`), desc(align.length), .by_group = TRUE) |>
  slice_head(n = 10) |>
  gt()
```

```{r echo=FALSE, message=FALSE}
renamed_files <- list.files(
  "~/project_data/downy/contigs-renamed",
  pattern = "\\.fasta\\.gz$",
  full.names = TRUE
)

clean_contigs <- function(file) {
  sample_name <- str_remove(basename(file), "\\.fasta\\.gz$")
  seqs <- readDNAStringSet(file)
  mito_ids <- mito_contigs |> filter(sample == sample_name) |> pull(seqid)

  clean_dir <- file.path("~/project_data/downy/contigs-renamed/cleaned")
  mito_dir <- file.path("~/project_data/downy/contigs-renamed/mitochondiral")
  dir.create(clean_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(mito_dir, recursive = TRUE, showWarnings = FALSE)

  writeXStringSet(
    seqs[!(names(seqs) %in% mito_ids)],
    file.path(clean_dir, basename(file)),
    compress = TRUE
  )
  writeXStringSet(
    seqs[names(seqs) %in% mito_ids],
    file.path(mito_dir, basename(file)),
    compress = TRUE
  )
}

walk(renamed_files, clean_contigs)
```
