---
title: "Gene Annotation and RNA-seq Support"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(gt)
library(here)

theme_set(theme_bw())
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

project_root <- path.expand(Sys.getenv("PROJECT_DATA", "~/project_data/downy"))
hardmask_dir <- file.path(project_root, "results/repeatmask-gene-prediction/focal/hardmasked")
rnaseq_dir <- file.path(project_root, "results/rnaseq-support")

isolates <- tribble(
  ~assembly,                            ~rnaseq,        ~label,                  ~plot_label,
  "Pseudoperonospora_cubensis_MSU1",   "Pcub_MSU1",    "*P. cubensis* MSU1",    "P. cubensis MSU1",
  "Pseudoperonospora_cubensis_SC1982", "Pcub_SC1982",  "*P. cubensis* SC1982",  "P. cubensis SC1982",
  "Pseudoperonospora_humuli_OR502AA",  "Phum_OR502AA", "*P. humuli* OR502AA",   "P. humuli OR502AA"
)
rna_support_threshold <- 2

read_repeatmasker_tbl <- function(assembly) {
  lines <- readLines(file.path(hardmask_dir, paste0(assembly, "_masked"), paste0(assembly, ".fasta.tbl")))
  bp_pct <- function(pattern) {
    line <- str_subset(lines, pattern)[1]
    pct <- str_match(line, "\\(\\s*([0-9.]+) %\\)|\\s([0-9.]+) %$")
    c(bp = as.numeric(str_match(line, "([0-9]+) bp")[, 2]), pct = as.numeric(coalesce(pct[, 2], pct[, 3])))
  }
  total <- as.numeric(str_match(str_subset(lines, "^total length:")[1], "([0-9]+) bp")[, 2])
  masked <- bp_pct("^bases masked:")
  tibble(
    assembly = assembly,
    `Assembly length (Mb)` = total / 1e6,
    `Masked bases (Mb)` = masked[["bp"]] / 1e6,
    `Masked (%)` = masked[["pct"]],
    `LTR elements (%)` = bp_pct("^   LTR elements:")[["pct"]],
    `Unclassified repeats (%)` = bp_pct("^Unclassified:")[["pct"]]
  )
}

# STAR mapping rate (unique + multi-mapped) of each RNA-seq library to its pathogen assembly.
read_star_mapping <- function(rnaseq) {
  logs <- list.files(file.path(rnaseq_dir, rnaseq, "star_salmon", "log"), pattern = "\\.Log\\.final\\.out$", full.names = TRUE)
  stopifnot(length(logs) > 0)
  map_dbl(logs, \(f) {
    x <- readLines(f)
    pct <- \(key) parse_number(str_subset(x, fixed(key))[1])
    pct("Uniquely mapped reads %") + pct("% of reads mapped to multiple loci")
  })
}

annotations <- read_tsv(here("data", "annotation_support.tsv"), show_col_types = FALSE) |>
  filter(isolate %in% isolates$assembly) |>
  rename(assembly = isolate) |>
  left_join(isolates, by = "assembly")

gene_level <- annotations |>
  group_by(assembly, label, plot_label, gene_id) |>
  summarise(
    has_function = any(source != "unannotated"),
    max_tpm = first(max_tpm),
    rna_supported = first(rna_supported_tpm_ge_2),
    .groups = "drop"
  )
# Every current gene model must have an RNA-seq TPM estimate.
stopifnot(nrow(gene_level) > 0, !anyNA(gene_level$max_tpm), all(is.finite(gene_level$max_tpm)),
          all(gene_level$rna_supported == (gene_level$max_tpm >= rna_support_threshold)))

mapping_range <- range(unlist(map(isolates$rnaseq, read_star_mapping)))

repeat_summary <- map_dfr(isolates$assembly, read_repeatmasker_tbl) |>
  left_join(isolates, by = "assembly") |>
  select(label, `Assembly length (Mb)`:`Unclassified repeats (%)`)

gene_summary <- gene_level |>
  group_by(label) |>
  summarise(
    `Predicted genes` = n(),
    `Functionally annotated` = sum(has_function),
    `Annotated (%)` = 100 * mean(has_function),
    `RNA-seq supported` = sum(rna_supported),
    `RNA-seq supported (%)` = 100 * mean(rna_supported),
    `Functional + RNA-seq support` = sum(has_function & rna_supported),
    `Hypothetical without RNA-seq support` = sum(!has_function & !rna_supported),
    .groups = "drop"
  )

support_summary <- gene_level |>
  mutate(support_class = factor(case_when(
    has_function & rna_supported ~ "Both",
    has_function ~ "Functional only",
    rna_supported ~ "RNA-seq only",
    TRUE ~ "Neither"
  ), levels = c("Both", "Functional only", "RNA-seq only", "Neither"))) |>
  count(plot_label, support_class, name = "genes") |>
  group_by(plot_label) |>
  mutate(percent = 100 * genes / sum(genes)) |>
  ungroup()
```

# Overview

Protein-coding genes were predicted with Helixer on the hard-masked assemblies of the three focal
isolates. Functional descriptions follow `workflow/functional-annotation/summarize-functions.R`:
an informative eggNOG-mapper description first, then the best InterProScan hit, then the best
DIAMOND BLASTP hit; proteins without any of these are kept as hypothetical proteins.
RNA-seq support is the maximum Salmon TPM across the infection time points and is treated as
transcript support for a gene model, not protein validation. Only
`r sprintf("%.1f-%.1f%%", mapping_range[1], mapping_range[2])` of RNA-seq reads mapped to the
pathogen assembly, so a gene is counted as supported at TPM >= `r rna_support_threshold` in at
least one time point to retain low-abundance transcript evidence.

# Repeat Masking

```{r repeat-summary}
repeat_summary |>
  gt() |>
  fmt_markdown(columns = label) |>
  cols_label(label = "Isolate") |>
  fmt_number(columns = -label, decimals = 1)
```

# Gene Models, Functional Annotation and RNA-seq Support

```{r gene-summary}
gene_summary |>
  gt() |>
  fmt_markdown(columns = label) |>
  cols_label(label = "Isolate") |>
  fmt_integer(columns = c(`Predicted genes`, `Functionally annotated`, `RNA-seq supported`,
                          `Functional + RNA-seq support`, `Hypothetical without RNA-seq support`)) |>
  fmt_number(columns = c(`Annotated (%)`, `RNA-seq supported (%)`), decimals = 1)
```

```{r support-plot, fig.width=7, fig.height=3.8}
support_plot <- support_summary |>
  mutate(plot_label = factor(plot_label, levels = rev(isolates$plot_label))) |>
  ggplot(aes(x = plot_label, y = percent, fill = support_class)) +
  geom_col(width = 0.75, position = position_stack(reverse = TRUE)) +
  coord_flip() +
  scale_y_continuous(
    breaks = seq(0, 100, 25),
    labels = \(x) paste0(x, "%"),
    limits = c(0, 102),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    x = NULL,
    y = "Predicted genes (%)",
    fill = NULL
  ) +
  scale_fill_manual(
    values = c(
      "Both" = "#0072B2",
      "Functional only" = "#56B4E9",
      "RNA-seq only" = "#E69F00",
      "Neither" = "#999999"
    )
  ) +
  guides(fill = guide_legend(ncol = 1, byrow = TRUE)) +
  theme_bw(base_size = 8) +
  theme(
    legend.position = "right",
    legend.key.size = unit(3.5, "mm"),
    legend.text = element_text(size = 8),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(3, 8, 3, 3)
  )

out <- here("figures", "gene-annotation-rnaseq-support")
ggsave(paste0(out, ".svg"), support_plot, device = svglite::svglite, width = 180, height = 75, units = "mm")
ggsave(paste0(out, ".pdf"), support_plot, device = grDevices::pdf, width = 180, height = 75, units = "mm", family = "Helvetica", useDingbats = FALSE)
ggsave(paste0(out, ".tiff"), support_plot, device = ragg::agg_tiff, width = 180, height = 75, units = "mm", dpi = 600, compression = "lzw")
ggsave(paste0(out, ".png"), support_plot, device = ragg::agg_png, width = 180, height = 75, units = "mm", dpi = 600)
support_plot
```
