---
title: "Pipeline result"
output: html_document
---

```{r echo=FALSE, message=FALSE}
library(gt)
library(data.table)
library(tidyverse)
```

```{r echo=FALSE, message=FALSE}
read.tb <- function(path, method) {
  files <- list.files(path, full.names = TRUE, pattern = 'stram.csv')
  tb <- map_dfr(files, function(x) {data.frame(method = method,
                                               sample = str_remove(basename(x), '_stram.csv'),
                                               fread(x))})
  return(tb)
}
```

# Pipeline

## hifiasm

Step 1, assemble the raw Pacbio read into contigs. Here we check the quality of the assembled genome.

Each isolates showed 100% completeness, but the duplications were high.

```{r echo=FALSE, message=FALSE}
read.tb('~/project_data/downy/hifiasm/compleasm/', 'hifiasm') |> gt() |> fmt_number()
```

## Decontamination

There could be many contamination reads in each isolates. Here, we use FCS-GX to clean the genome.

```{r echo=FALSE, message=FALSE}
read.tb('~/project_data/downy/hifiasm/fcs-gx/compleasm/', 'FCS') |> gt() |> fmt_number()
```

We notice there are some contigs that are not classified by fcs. Here, we check the fcs cleaned data with kraken2

```{r echo=FALSE, message=FALSE}
read.tb('~/project_data/downy/hifiasm/fcs-gx/kraken2/compleasm/', 'Kraken2') |> gt() |> fmt_number()
```

## Purge_dups

The duplication rates are still too high. Here we can try to use purge_dups software. However, purge_dups sometimes fails on setting threshold parameters. Therefore, we set them manually. 

Purge_dups parameters:

MSU1: -l 80 -m 165 -u 360

Phumuli: -l 80 -m 220 -u 400

SC1982: -l 7 -m 65 -u 180

```{r echo=FALSE, message=FALSE}
read.tb('~/project_data/downy/hifiasm/fcs-gx/kraken2/purge_dups/compleasm/', 'Purge_dups') |> gt() |> fmt_number()
```

# Synteny with ragtag

```{r echo=FALSE, message=FALSE}
read.tb('~/project_data/downy/hifiasm/fcs-gx/kraken2/purge_dups/rag_tag/compleasm/', 'Rag_tag') |> gt() |> fmt_number()
```

# Trace Plots

```{r echo=FALSE}
df <- bind_rows(read.tb('~/project_data/downy/hifiasm/compleasm/', 'hifiasm'), 
                read.tb('~/project_data/downy/hifiasm/fcs-gx/compleasm/', 'FCS'),
                read.tb('~/project_data/downy/hifiasm/fcs-gx/kraken2/compleasm/', 'Kraken2'),
                read.tb('~/project_data/downy/hifiasm/fcs-gx/kraken2/purge_dups/compleasm/', 'Purge_dups'),
                read.tb('~/project_data/downy/hifiasm/fcs-gx/kraken2/purge_dups/rag_tag/compleasm/', 'Rag_tag'))
df$method <- factor(df$method, levels = c('hifiasm', 'FCS', 'Kraken2', 'Purge_dups', 'Rag_tag'))
```

```{r echo=FALSE}
ggplot(df, aes(x = method, y = X..contigs, color = sample, group = sample)) +
  geom_point() + geom_line() + ylab('Number of contigs')
```

```{r echo=FALSE}
ggplot(df, aes(x = method, y = Total.length, color = sample, group = sample)) +
  geom_point() + geom_line()
```

```{r echo=FALSE}
ggplot(df, aes(x = method, y = N50, color = sample, group = sample)) +
  geom_point() + geom_line()
```