---
title: "Pipeline result"
output: html_document
---

```{r echo=FALSE, message=FALSE}
library(gt)
library(data.table)
library(tidyverse)
```

```{r echo=FALSE, message=FALSE}
read.tb <- function(path, method) {
  files <- list.files(path, full.names = TRUE, pattern = 'stram.csv')
  tb <- map_dfr(files, function(x) {data.frame(method = method,
                                               sample = str_remove(basename(x), '_stram.csv'),
                                               fread(x))})
  return(tb)
}
```


```{r echo=FALSE, message=FALSE, include=FALSE}
result <- bind_rows(read.tb('~/project_data/downy/hifiasm/compleasm/', 's.13.a'),
                    read.tb('~/project_data/downy/hifiasm-10kb/compleasm/', 's.13.b'),
                    read.tb('~/project_data/downy/hifiasm/purge_dups/compleasm/', 's.13.c'),
                    read.tb('~/project_data/downy/hifiasm-10kb/purge_dups/compleasm/', 's.13.d'),
                    read.tb('~/project_data/downy/hifiasm/fcs-gx/purge_dups/compleasm/', 's.13.e'),
                    read.tb('~/project_data/downy/hifiasm-10kb/fcs-gx/purge_dups/compleasm/', 's.13.f'))

result %>% group_by(sample) %>% gt()
```

# Pipeline

## hifiasm

Step 1, assemble the raw Pacbio read into contigs. Here we check the quality of the assembled genome.

Each isolates showed 100% completeness, but the duplications were high.

```{r}
read.tb('~/project_data/downy/hifiasm/compleasm/', 'hifiasm') |> gt()
```

## Decontamination

There could be many contamination reads in each isolates. Here, we use FCS-GX to clean the genome.

```{r}
read.tb('~/project_data/downy/hifiasm/fcs-gx/compleasm/', 'FCS') |> gt()
```

## Purge_dups

The duplication rates are still too high. Here we can try to use purge_dups software. However, purge_dups sometimes fails on setting threshold parameters. Therefore, we set them manually. 

Purge_dups parameters:

MSU1: -l 80 -m 165 -u 360

Phumuli: -l 80 -m 220 -u 400

SC1982: -l 7 -m 65 -u 180

```{r echo=FALSE, message=FALSE}
read.tb('~/project_data/downy/hifiasm/fcs-gx/purge_dups/compleasm/', 'Purge_dups') |> gt()
```


**we can set more aggressive parameters**

MSU1: -l 80 -m 220 -u 360

Phumuli: -l 80 -m 220 -u 400

SC1982: -l 7 -m 120 -u 180

```{r echo=FALSE, message=FALSE}
read.tb('~/project_data/downy/hifiasm/fcs-gx/purge_dups-old/compleasm/', 'Purge_dups optimized') |> gt()
```


