---
title: "Read Distribution and Taxonomy Analysis"
output: html_document
---

```{r setup, echo=FALSE, message=FALSE}
library(tidyverse)
library(data.table)
library(here)
library(patchwork)
theme_set(theme_bw(base_size = 7, base_family = "Helvetica"))
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

project_root <- path.expand(Sys.getenv("PROJECT_DATA", "~/project_data/downy"))
taxonomy_dir <- file.path(project_root, "results/read-filtering-screening/taxonomy")
coverage_dir <- file.path(project_root, "results/read-filtering-screening/coverage-gc")

# One row per HiFi library; `sample` is the file basename used by the read-filtering outputs.
isolates <- tribble(
  ~isolate,   ~sample,                 ~label,
  "MSU1",     "Quesada_SQIIe_MSU1",    "P. cubensis (MSU1)",
  "SC1982",   "Quesada_SQIIe_SC1982",  "P. cubensis (SC1982)",
  "OR502AA",  "Quesada_SQIIe_Phumuli", "P. humuli (OR502AA)",
  "UA202013", "UA202013",              "P. effusa (UA202013)"
) |>
  mutate(label = factor(label, levels = label))

taxon_colors <- c(
  Oomycota = "#000000", Streptophyta = "#0072B2", Pseudomonadota = "#D55E00",
  Bacteroidota = "#009E73", Ascomycota = "#CC79A7", Chordata = "#E69F00",
  unclassified = "#56B4E9", Other = "#999999"
)

read_kreport <- function(file) {
  report <- read.delim(
    file, header = FALSE, sep = "\t", quote = "", comment.char = "",
    strip.white = FALSE,
    col.names = c("percentage", "cladeReads", "taxonReads", "taxRank", "taxID", "name")
  )
  report$depth <- (nchar(report$name) - nchar(sub("^\\s+", "", report$name))) / 2
  report$name <- paste0(tolower(report$taxRank), "_", trimws(report$name))
  report$percentage <- round(report$cladeReads / sum(report$taxonReads), 6) * 100
  report$P <- NA_character_
  phylum <- NA_character_
  phylum_depth <- -1
  for (i in seq_len(nrow(report))) {
    if (report$depth[i] <= phylum_depth) phylum <- NA_character_
    if (report$taxRank[i] == "P") {
      phylum <- report$name[i]
      phylum_depth <- report$depth[i]
    }
    report$P[i] <- phylum
  }
  report
}
clean_taxon <- function(x) str_remove(x, "^[a-z]_")

oomycete_ids <- read_lines(here("data", "oomycete_taxids.txt"))  # taxids under Oomycota (4762)
```

```{r load-kraken-data}
reads <- pmap_dfr(isolates, \(isolate, sample, label) {
  fread(file.path(taxonomy_dir, paste0(sample, ".kraken")), drop = 5, showProgress = FALSE,
        col.names = c("status", "id", "tax", "length")) |>
    mutate(isolate = isolate, label = label)
}) |>
  mutate(is_oomycota = tax %in% oomycete_ids)

kreports <- pmap_dfr(isolates, \(isolate, sample, label) {
  read_kreport(file.path(taxonomy_dir, paste0(sample, ".kreport"))) |>
    mutate(isolate = isolate, label = label)
})
stopifnot(setequal(unique(reads$isolate), isolates$isolate), setequal(unique(kreports$isolate), isolates$isolate))

read_totals <- reads |>
  group_by(label, isolate) |>
  summarise(read_count = n(), total_bases = sum(as.numeric(length)), .groups = "drop") |>
  mutate(total_gbp = total_bases / 1e9)
write_tsv(select(read_totals, isolate, read_count, total_bases, total_gbp), here("data", "figure1_read_totals.tsv"))
library_labels <- with(read_totals, setNames(
  sprintf("%s\n%s reads; %.2f Gbp total", label, scales::comma(read_count), total_gbp), label
))
```

# Read length distribution of the PacBio HiFi reads

```{r read-length-distribution, fig.width=7.2, fig.height=5.9}
read_dist <- reads |>
  mutate(is_oomycota = if_else(is_oomycota, "Oomycota-assigned", "Non-Oomycota")) |>
  ggplot() +
  geom_histogram(aes(x = length, fill = is_oomycota), binwidth = 1000) +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_fill_manual(values = c("Oomycota-assigned" = "#000000", "Non-Oomycota" = "#999999")) +
  labs(x = "Read length (bp)", y = "Read count", fill = NULL) +
  facet_wrap(~label, ncol = 1, scales = "free_y")
out <- here("figures", "supp-read-length-distribution")
ggsave(paste0(out, ".svg"), read_dist, device = svglite::svglite, width = 180, height = 100, units = "mm")
ggsave(paste0(out, ".pdf"), read_dist, device = grDevices::pdf, width = 180, height = 100, units = "mm", family = "Helvetica", useDingbats = FALSE)
ggsave(paste0(out, ".tiff"), read_dist, device = ragg::agg_tiff, width = 180, height = 100, units = "mm", dpi = 600, compression = "lzw")
ggsave(paste0(out, ".png"), read_dist, device = ragg::agg_png, width = 180, height = 100, units = "mm", dpi = 600)
read_dist
```

# Taxonomic composition and read GC/coverage

The composition panel uses Kraken2 (PlusPFP) phylum-level assignments; phyla below 5% of a
library's reads are collapsed to "Other". The GC/coverage panel plots per-read GC content
(seqkit `fx2tab`) against per-read median 21-mer self-coverage (KAT `sect`) for the
adapter-filtered HiFi reads, coloured by the same phylum categories. Coverage is a proxy for
how often a molecule was sampled, not a per-base assembly coverage. Category summaries are
written to `data/blobplot-category-summary-by-isolate.tsv`.

```{r taxonomy-panel}
tax_bar <- kreports |>
  filter(taxRank %in% c("U", "P")) |>
  mutate(taxon = if_else(percentage < 5, "Other", clean_taxon(name))) |>
  ggplot() +
  geom_col(aes(x = percentage, y = label, fill = taxon), position = "fill") +
  scale_x_continuous(labels = scales::percent) +
  scale_y_discrete(labels = library_labels) +
  scale_fill_manual(values = taxon_colors, guide = "none") +
  labs(x = "Reads (%)", y = NULL, fill = "Taxon")
```

```{r coverage-gc-data}
# Resolve each read's Kraken2 taxID to its phylum with the PlusPFP taxonomy.
taxonkit <- Sys.getenv("TAXONKIT_BIN", file.path(Sys.getenv("HOME"), "miniforge3/envs/downy/bin/taxonkit"))
taxdump <- file.path(path.expand(Sys.getenv("DB_ROOT", "~/db")), "kraken2/PlusPFP")
taxids <- setdiff(unique(as.character(reads$tax)), c(NA, "0"))
taxid_file <- tempfile(fileext = ".txt")
writeLines(taxids, taxid_file)
phylum_lines <- system2(taxonkit, c("reformat", "-I", "1", "-f", shQuote("{p}"), "--data-dir", taxdump, taxid_file),
                        stdout = TRUE, stderr = FALSE)
unlink(taxid_file)
taxid_phylum <- fread(text = phylum_lines, header = FALSE, sep = "\t", col.names = c("taxID", "phylum"),
                      colClasses = "character", showProgress = FALSE) |>
  mutate(phylum = na_if(phylum, ""))
stopifnot(nrow(taxid_phylum) == length(taxids))

plot_data <- pmap_dfr(isolates, \(isolate, sample, label) {
  phylum_category <- kreports |>
    filter(isolate == !!isolate, taxRank == "P") |>
    transmute(phylum = clean_taxon(name), category = if_else(percentage < 5, "Other", phylum))
  gc <- fread(file.path(coverage_dir, paste0(sample, "-gc.tsv")), header = FALSE,
              col.names = c("read_id", "gc_pct"), showProgress = FALSE) |>
    mutate(read_id = word(read_id, 1))
  cov <- fread(file.path(coverage_dir, paste0(sample, "-sect-stats.tsv")),
               select = c("seq_name", "median"), showProgress = FALSE) |>
    transmute(read_id = word(seq_name, 1), cov_median = median)
  reads |>
    filter(isolate == !!isolate) |>
    transmute(read_id = id, taxID = as.character(tax)) |>
    left_join(taxid_phylum, by = "taxID") |>
    left_join(phylum_category, by = "phylum") |>
    mutate(category = case_when(taxID == "0" ~ "unclassified", is.na(category) ~ "Other", TRUE ~ category)) |>
    inner_join(gc, by = "read_id") |>
    inner_join(cov, by = "read_id") |>
    transmute(isolate = isolate, label = label, category, gc_pct, cov_median)
})

category_summary <- plot_data |>
  group_by(label, isolate, category) |>
  summarise(n_reads = n(), gc_mean = mean(gc_pct), cov_mean_of_medians = mean(cov_median), .groups = "drop")
write_tsv(select(category_summary, -label), here("data", "blobplot-category-summary-by-isolate.tsv"))
```

```{r read-qc-profile, fig.width=7.09, fig.height=6.5}
blob_plot <- plot_data |>
  ggplot(aes(x = gc_pct, y = cov_median, color = category)) +
  ggrastr::geom_point_rast(alpha = 0.1, size = 0.25, raster.dpi = 600, dev = "ragg") +
  scale_y_log10(labels = scales::label_comma()) +
  scale_color_manual(values = taxon_colors, labels = \(x) ifelse(x == "unclassified", "Unclassified", x)) +
  facet_wrap(~label, nrow = 1) +
  guides(color = guide_legend(nrow = 2, override.aes = list(alpha = 1, size = 2.5))) +
  labs(x = "GC content (%)", y = "Median 21-mer coverage", color = "Taxon")

read_qc_profile <- tax_bar / blob_plot +
  plot_layout(ncol = 1, heights = c(1, 0.85)) +
  plot_annotation(tag_levels = "A") &
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 9),
    legend.key.size = grid::unit(5, "mm")
  )

out <- here("figures", "read-qc-profile")
ggsave(paste0(out, ".svg"), read_qc_profile, device = svglite::svglite, width = 180, height = 165, units = "mm")
ggsave(paste0(out, ".pdf"), read_qc_profile, device = grDevices::pdf, width = 180, height = 165, units = "mm", family = "Helvetica", useDingbats = FALSE)
ggsave(paste0(out, ".tiff"), read_qc_profile, device = ragg::agg_tiff, width = 180, height = 165, units = "mm", dpi = 600, compression = "lzw")
ggsave(paste0(out, ".png"), read_qc_profile, device = ragg::agg_png, width = 180, height = 165, units = "mm", dpi = 600)
read_qc_profile
```

```{r category-summary}
category_summary |>
  arrange(label, desc(n_reads)) |>
  select(Isolate = label, Category = category, Reads = n_reads, `Mean GC (%)` = gc_mean,
         `Mean median coverage` = cov_mean_of_medians) |>
  knitr::kable(digits = 1, format.args = list(big.mark = ","))
```

## Finding

The full-read profiles reveal taxon-associated GC and coverage patterns. `Pseudomonadota` is
concentrated at relatively high GC content and low 21-mer coverage, while `Bacteroidota` tends
toward lower GC content and low coverage. `Oomycota` often has higher coverage than these
bacterial groups but overlaps substantially with `Streptophyta` and unclassified reads.
GC content and coverage can therefore help identify some non-target populations, but a simple
threshold on these axes would not cleanly isolate Oomycota-assigned reads. Reads unclassified
by Kraken2 should not be assumed to be contaminants.
