---
title: "Read Distribution and Taxonomy Analysis"
output: html_document
---

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(data.table)
library(taxizedb)
library(gt)
library(here)
library(pavian)
theme_set(theme_bw())
```

# Read distribution of the PacBio HIFI reads

```{r echo=FALSE}
# child ids under oomycota: 4762
child.id <- c(taxizedb::downstream(4762)$`4762`$childtaxa_id, '4762')
files <- list.files('~/project_data/downy/Kraken2',
                    full.names = TRUE, recursive = TRUE, pattern = '.kraken$')
df <- map_dfr(files, function(file) {
  data <- fread(file, drop = 5, showProgress = FALSE)
  colnames(data) <- c('status', 'id', 'tax', 'length')
  data <- data %>% mutate(file = str_remove(basename(file), '.kraken'))
}) %>% mutate(is.oomycota = (tax %in% child.id))

ggplot(df) +
  geom_histogram(aes(x = length, fill = is.oomycota), binwidth = 1000) +
  scale_y_continuous(labels = scales::label_comma()) +
  facet_wrap(~file, ncol = 1, scales = 'free_y')
```

# Taxonomy abundance of PacBio raw reads

```{r echo=FALSE}
files <- list.files('~/project_data/downy/Kraken2/raw',
                    full.names = TRUE, recursive = TRUE, pattern = '.kreport$')
df.rpt <- map_dfr(files, function(file) {
  read_report2(file, add_taxRank_columns = TRUE) %>%
    mutate(file = str_remove(basename(file), '.fastq.gz.kreport'))})
```

Is there any interesting species other than oomycota? Here, we are showing top 10 species level taxa with highest percentage for each isolates.

```{r echo=FALSE}
df.rpt %>% 
  filter(taxRank %in% c('U', 'S')) %>% 
  select(percentage, P, S, file) %>% 
  group_by(file) %>% 
  slice_max(order_by = percentage, n = 10) %>% 
  gt(row_group_as_column = T) %>% 
  fmt_number()
```

# Taxonomy of contigs before using FCS

Here we are looking at top 10 species level taxa with highest percentage for each isolates.

```{r echo=FALSE}
files <- list.files('~/project_data/downy/Kraken2/asm',
                    full.names = TRUE, recursive = TRUE, pattern = '.kreport$')
map_dfr(files, function(file) {
  read_report2(file, add_taxRank_columns = TRUE) %>%
  mutate(file = str_remove(basename(file), '.fasta.gz.kreport'))}) %>% 
  filter(taxRank %in% c('U', 'S')) %>% 
  select(percentage, P, S, file) %>% 
  group_by(file) %>% 
  slice_max(order_by = percentage, n = 10) %>% 
  gt(row_group_as_column = T) %>% 
  fmt_number()
```

# Taxonomy of contigs in final assembly

We are looking at top 10 phylum level taxa with highest percentage for each isolates to check if most of the contigs belong to oomycete. 

In isolate SC1982, the taxonomy is still diverse, and only a small part of contigs classified as oomycete. This might because this isolate has many non-coding regions.

```{r echo=FALSE}
files <- list.files('~/project_data/downy/Kraken2/final',
                    full.names = TRUE, recursive = TRUE, pattern = '.kreport$')
map_dfr(files, function(file) {
  read_report2(file, add_taxRank_columns = TRUE) %>%
  mutate(file = str_remove(basename(file), '.fasta.gz.kreport'))}) %>% 
  filter(taxRank %in% c('U', 'P')) %>% 
  select(percentage, P, file) %>% 
  group_by(file) %>% 
  slice_max(order_by = percentage, n = 10) %>% 
  gt(row_group_as_column = T) %>% 
  fmt_number()
```

