---
title: "Read Distribution and Taxonomy Analysis"
output: html_document
---

```{r echo=FALSE, message=FALSE}
source(here::here("analysis", "lib", "paths.R"))
library(tidyverse)
library(data.table)
library(gt)
library(here)
library(patchwork)
source(here('analysis', 'lib', 'figure-utils.R'))
# read_report2() (Kraken report parser) vendored from pavian rather than depending on the
# full package - see analysis/lib/read-kraken-report.R for why.
source(here('analysis', 'lib', 'read-kraken-report.R'))
theme_set(theme_pub())
# child ids under oomycota: 4762
oomycete.ids <- read_lines(here('data', 'oomycete_taxids.txt'))
sample.labels <- c(
  # Kraken2 taxonomy files for the public P. effusa reads are named after the isolate
  # accession (p_effusa.kraken/.kreport), not "UA202013" - matching the recode key to the
  # actual file basename is what makes this facet resolve instead of showing up as NA.
  "Quesada_SQIIe_MSU1" = "P. cubensis (MSU1)",
  "Quesada_SQIIe_SC1982" = "P. cubensis (SC1982)",
  "Quesada_SQIIe_Phumuli" = "P. humuli (OR502AA)",
  "p_effusa" = "P. effusa (UA202013)"
)
label_isolate <- function(data) {
  mutate(data, Isolate = factor(recode(file, !!!sample.labels), levels = sample.labels))
}
clean_taxon <- function(x) str_remove(x, "^[a-z]_")
isolate.labels <- c(
  "MSU1" = "P. cubensis (MSU1)",
  "SC1982" = "P. cubensis (SC1982)",
  "Phumuli" = "P. humuli (OR502AA)",
  "UA202013" = "P. effusa (UA202013)"
)
isolates <- tibble::tribble(
  ~isolate,   ~taxonomy_sample,        ~coverage_sample,
  "MSU1",     "Quesada_SQIIe_MSU1",    "Quesada_SQIIe_MSU1",
  "SC1982",   "Quesada_SQIIe_SC1982",  "Quesada_SQIIe_SC1982",
  "Phumuli",  "Quesada_SQIIe_Phumuli", "Quesada_SQIIe_Phumuli",
  "UA202013", "p_effusa",              "p_effusa"
)
```

```{r load-kraken-data, echo=FALSE, message=FALSE}
taxonomy.files <- list.files(
  project_path("results/read-filtering-screening/taxonomy"),
  full.names = TRUE,
  recursive = TRUE,
  pattern = "\\.kraken$"
)
df <- map_dfr(taxonomy.files, \(file) {
  fread(file, drop = 5, showProgress = FALSE, col.names = c("status", "id", "tax", "length")) |>
    mutate(file = tools::file_path_sans_ext(basename(file)))
}) |>
  mutate(is.oomycota = tax %in% oomycete.ids) |>
  label_isolate()

report.files <- list.files(
  project_path("results/read-filtering-screening/taxonomy"),
  full.names = TRUE,
  recursive = TRUE,
  pattern = "\\.kreport$"
)
df.rpt <- map_dfr(report.files, \(file) {
  read_report2(file, add_taxRank_columns = TRUE) |>
    mutate(file = str_remove(basename(file), "\\.kreport$"))
}) |>
  label_isolate()
```

# Read distribution of the PacBio HIFI reads

```{r read-length-distribution, echo=FALSE, fig.width=7.2, fig.height=5.9}
library.labels <- df |>
  group_by(Isolate) |>
  summarise(reads = n(), bases = sum(length), .groups = "drop") |>
  transmute(
    Isolate,
    label = sprintf(
      "%s\n%s reads; %.2f Gbp total",
      Isolate,
      scales::comma(reads),
      bases / 1e9
    )
  ) |>
  deframe()

read.dist <- df |>
  mutate(is.oomycota = if_else(is.oomycota, "Oomycota-assigned", "Non-Oomycota")) |>
  ggplot() +
  geom_histogram(aes(x = length, fill = is.oomycota), binwidth = 1000) +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_fill_manual(values = c("Oomycota-assigned" = unname(taxon_palette["Oomycota"]), "Non-Oomycota" = unname(taxon_palette["Other"]))) +
  labs(x = "Read length (bp)", y = "Read count", fill = NULL) +
  facet_wrap(~Isolate, ncol = 1, scales = 'free_y')
save_pub_r(read.dist, 'supp-read-length-distribution', width_mm = 180, height_mm = 100)
read.dist
```

```{r echo=FALSE}
df.rpt |> 
  group_by(Isolate) |> 
  filter(name == "p_Oomycota") |> 
  select(percentage, cladeReads, name, Isolate)
```


```{r}
df.rpt |> 
  group_by(Isolate) |> 
  filter(name == "g_Homo") |> 
  select(percentage, cladeReads, name, Isolate)
```

```{r echo=FALSE}
df.rpt |> 
  group_by(Isolate) |> 
  filter(name %in% c("f_Cannabaceae", "f_Cucurbitaceae", "f_Amaranthaceae")) |> 
  select(percentage, cladeReads, name, Isolate) |> 
  gt()
```


```{r read-qc-taxonomy-panel, echo=FALSE}
tax.bar <- df.rpt |> 
  group_by(Isolate) |> 
  filter(taxRank %in% c('U', 'P')) |> 
  mutate(taxon = if_else(percentage < 5, 'Other', clean_taxon(name))) |>
  ggplot() +
  geom_col(aes(x = percentage, y = Isolate, fill = taxon), position = 'fill') +
  scale_x_continuous(labels = scales::percent) +
  scale_y_discrete(labels = library.labels) +
  scale_fill_manual(values = taxon_palette) +
  labs(x = "Reads (%)", y = NULL, fill = "Taxon")
```

```{r}
df.rpt |> 
  filter(!(taxID %in% oomycete.ids), P == 'p_Pseudomonadota', taxRank == 'G') |> 
  group_by(Isolate) |> 
  arrange(desc(percentage)) |> 
  slice_head(n = 10) |> 
  gt()
```

# Can GC content and coverage separate target Oomycota reads from contamination?

Per-read GC content (seqkit `fx2tab`) and per-read 21-mer self-coverage (KAT `sect`) for the
adapter-filtered HiFi reads of all four isolates, colored by the same Kraken2 phylum-level taxon
call used for the read-composition panel above (`taxRank %in% c('U','P')`, phyla below 5% of a
library's reads collapsed to "Other"). This is the read-level analogue of a blobtools blob plot:
assembly-level GC/coverage is not informative here because most of these libraries are
pre-assembly and dominated by non-target DNA (see the taxonomic composition above), so the
question is asked directly of the reads.

Coverage is per-read median 21-mer depth from KAT `sect` in self-comparison mode (each read's
own k-mers scored against the k-mer spectrum of the full read set) - a proxy for how many times
that piece of the underlying genome/molecule was sampled, not a per-base assembly coverage.
Points are a stratified random subsample (up to 3,000 reads per isolate x category) for
plotting; category summary statistics (`data/blobplot-category-summary-by-isolate.tsv`) are
computed from the full read sets.

```{r coverage-gc-data, echo=FALSE, message=FALSE, warning=FALSE}
# Resolve every read's Kraken2 taxID to its phylum using the same PlusPFP taxonomy dump.
# Keeping this beside the taxonomy profile makes both panels use one categorization rule.
TAXONKIT_BIN <- Sys.getenv(
  "TAXONKIT_BIN",
  file.path(Sys.getenv("HOME"), "miniforge3/envs/downy/bin/taxonkit")
)
TAXDUMP_DIR <- db_path("kraken2/PlusPFP")
all.taxids <- unique(as.character(df$tax))
all.taxids <- all.taxids[!is.na(all.taxids) & all.taxids != "0"]

taxid.file <- tempfile(fileext = ".txt")
writeLines(all.taxids, taxid.file)
phylum.lines <- system2(
  TAXONKIT_BIN,
  c("reformat", "-I", "1", "-f", shQuote("{p}"), "--data-dir", TAXDUMP_DIR, taxid.file),
  stdout = TRUE,
  stderr = FALSE
)
unlink(taxid.file)
taxid.phylum <- fread(
  text = phylum.lines,
  header = FALSE,
  sep = "\t",
  col.names = c("taxID", "phylum"),
  colClasses = "character",
  showProgress = FALSE
) |>
  mutate(phylum = na_if(phylum, ""))

set.seed(1)
max.per.category <- 3000
subsample.list <- list()
summary.list <- list()

for (i in seq_len(nrow(isolates))) {
  isolate <- isolates$isolate[i]
  taxonomy.sample <- isolates$taxonomy_sample[i]
  coverage.sample <- isolates$coverage_sample[i]

  kreport.file <- project_path(
    "results/read-filtering-screening/taxonomy",
    paste0(taxonomy.sample, ".kreport")
  )
  gc.file <- project_path(
    "results/read-filtering-screening/coverage-gc",
    paste0(coverage.sample, "-gc.tsv")
  )
  sect.file <- project_path(
    "results/read-filtering-screening/coverage-gc",
    paste0(coverage.sample, "-sect-stats.tsv")
  )

  phylum.category <- read_report2(kreport.file, add_taxRank_columns = TRUE) |>
    filter(taxRank == "P") |>
    transmute(
      phylum = clean_taxon(name),
      category = if_else(percentage < 5, "Other", phylum)
    )

  reads <- df |>
    filter(file == taxonomy.sample) |>
    transmute(read_id = id, taxID = as.character(tax)) |>
    left_join(taxid.phylum, by = "taxID") |>
    left_join(phylum.category, by = "phylum") |>
    mutate(category = case_when(
      taxID == "0" ~ "unclassified",
      is.na(category) ~ "Other",
      TRUE ~ category
    )) |>
    select(read_id, category)

  gc <- fread(
    gc.file,
    header = FALSE,
    col.names = c("read_id", "gc_pct"),
    showProgress = FALSE
  ) |>
    mutate(read_id = word(read_id, 1))

  cov <- fread(sect.file, select = c("seq_name", "median"), showProgress = FALSE) |>
    rename(read_id = seq_name, cov_median = median) |>
    mutate(read_id = word(read_id, 1))

  joined <- reads |>
    inner_join(gc, by = "read_id") |>
    inner_join(cov, by = "read_id") |>
    mutate(isolate = isolate)

  summary.list[[isolate]] <- joined |>
    group_by(isolate, category) |>
    summarise(
      n_reads = n(),
      gc_mean = mean(gc_pct),
      cov_mean_of_medians = mean(cov_median),
      .groups = "drop"
    )

  subsample.list[[isolate]] <- joined |>
    group_by(category) |>
    group_modify(~ slice_sample(.x, n = min(max.per.category, nrow(.x)))) |>
    ungroup() |>
    select(isolate, read_id, category, gc_pct, cov_median)
}

plot_data <- bind_rows(subsample.list) |>
  mutate(Isolate = factor(recode(isolate, !!!isolate.labels), levels = isolate.labels))
summary.tbl <- bind_rows(summary.list)
write_tsv(plot_data |> select(-Isolate), here("data", "blobplot-read-coverage-gc-subsample.tsv"))
write_tsv(summary.tbl, here("data", "blobplot-category-summary-by-isolate.tsv"))
```

```{r read-qc-profile, fig.width=7.09, fig.height=6.5, message=FALSE}

blob.plot <- plot_data |>
  ggplot(aes(x = gc_pct, y = cov_median, color = category)) +
  geom_point(alpha = 0.25, size = 0.4) +
  scale_y_log10(labels = scales::label_comma()) +
  scale_color_manual(values = taxon_palette) +
  facet_wrap(~Isolate, nrow = 1) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.2))) +
  labs(x = "GC content (%)", y = "Median 21-mer coverage", color = "Taxon")

read.qc.profile <- tax.bar / blob.plot +
  plot_layout(ncol = 1, heights = c(1, 0.85), guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")

save_pub_r(read.qc.profile, "read-qc-profile", width_mm = 180, height_mm = 165)
read.qc.profile
```

```{r echo=FALSE}
summary.tbl |>
  mutate(Isolate = factor(recode(isolate, !!!isolate.labels), levels = isolate.labels)) |>
  arrange(Isolate, desc(n_reads)) |>
  select(Isolate, category, n_reads, gc_mean, cov_mean_of_medians)
```

## Finding

Coverage separates some categories better than GC content does - e.g. `Pseudomonadota` and
`Bacteroidota` sit at consistently low coverage relative to the rest of each library - but
**neither axis, nor the combination, cleanly separates target Oomycota reads from
contamination**. `Oomycota` does not form a single distinct cluster in any of the four panels:
it spans essentially the whole coverage range visible in each isolate, overlapping heavily with
`Streptophyta` (host), `Other` and `unclassified`. Conversely, every visually distinct
cloud/band in the plot is a mixture of taxa rather than a pure category. GC content alone is
close to uninformative for this separation (categories overlap almost completely on the x-axis
in all four isolates). A blob-plot-style GC/coverage view, at either the assembly or the read
level, is not sufficient here to distinguish target genome reads from contaminant or host reads.
