---
title: "report-0325"
output: html_document
---

```{r}
library(tidyverse)
library(gt)
library(scales)
format_large_numbers <- function(x) {
  ifelse(grepl("^\\d+(\\.\\d+)?$", x) & as.numeric(x) > 1000, 
         comma(as.numeric(x)), x)
}
map.result <- function(file, suffix) {
    df <- read.csv(file) %>% 
      column_to_rownames('Metric')
    # df$Value <- format_large_numbers(df$Value)
    sample <- basename(file) %>% str_replace(suffix, '')
    colnames(df) <- sample
    return(df)
}

```

read in data

```{r}
fcs.cleaning.euk <- list.files('~/project_data/downy/fcs-genome-cleaning/', full.names = TRUE, recursive = TRUE, pattern = 'fcs_cleaned_euk') %>% 
  map_dfc(function(x) map.result(file = x, suffix = '.csv'))

fcs.cleaning.stram <- list.files('~/project_data/downy/fcs-genome-cleaning/', full.names = TRUE, recursive = TRUE, pattern = 'fcs_cleaned_stram') %>% 
  map_dfc(function(x) map.result(file = x, suffix = '.csv'))

kraken.cleaning.euk <- list.files('~/project_data/downy/fcs-genome-cleaning/', full.names = TRUE, recursive = TRUE, pattern = 'kraken_cleaned_euk') %>% 
  map_dfc(function(x) map.result(file = x, suffix = '.csv'))

kraken.cleaning.stram <- list.files('~/project_data/downy/fcs-genome-cleaning/', full.names = TRUE, recursive = TRUE, pattern = 'kraken_cleaned_stram') %>% 
  map_dfc(function(x) map.result(file = x, suffix = '.csv'))
```

# Pipeline

1. Raw reads are assembled into contigs by `metaMDBG`

2. use `FCS-GX` to clean the contigs. Contamination reads are removed (action labeled as `REVIEW` are kept). Contig with length < 5000bp are removed. Database was retrieved from https://ftp.ncbi.nlm.nih.gov/genomes/TOOLS/FCS/database/latest/all.manifest. Compleasm (BUSCO) and QUAST analyses were preformed.

3. use `Kraken2` to re-clean the result of `FCS-GX`. Only extract the contigs that are classified under oomycetes. The database contain human, fungi, bacteria, cucumber, hops, and candidate oomycete genome were used. Compleasm (BUSCO) and QUAST analyses were preformed. confidence score is 0.5

# Compare FCS and Kraken-re-cleaned result

Eukaryota linage

```{r}
data.frame(fcs.cleaning.euk, kraken.cleaning.euk) %>%
  select(sort(names(.))) %>%
  mutate(across(everything(), format_large_numbers))
```

Stramenopiles linage

```{r}
data.frame(fcs.cleaning.stram, kraken.cleaning.stram) %>%
  select(sort(names(.))) %>%
  mutate(across(everything(), format_large_numbers)) %>% 
  head(7)
```

# Karken2: large database v.s. small database

In the previous section, we used a large database in Kraken2, which could be unnecessary. Here we compare the result when using a small database (cucumber, hops, and candidate oomycete genome) and the large database (cucumber, hops, candidate oomycete genome, human, fungi, bacteria).

```{r}
kraken.cleaning.euk.small <- list.files('~/project_data/downy/fcs-genome-cleaning-small/', full.names = TRUE, recursive = TRUE, pattern = 'kraken_cleaned_euk') %>% 
  map_dfc(function(x) map.result(file = x, suffix = '.csv'))
kraken.cleaning.stram.small <- list.files('~/project_data/downy/fcs-genome-cleaning-small/', full.names = TRUE, recursive = TRUE, pattern = 'kraken_cleaned_stram') %>% 
  map_dfc(function(x) map.result(file = x, suffix = '.csv'))

colnames(kraken.cleaning.euk.small) <- colnames(kraken.cleaning.euk.small) %>% str_replace('_cleaned_euk','_small')
colnames(kraken.cleaning.stram.small) <- colnames(kraken.cleaning.stram.small) %>% str_replace('_cleaned_stram','_small')
```

```{r}
data.frame(kraken.cleaning.euk, kraken.cleaning.euk.small) %>%
  select(sort(names(.))) %>%
  mutate(across(everything(), format_large_numbers))
```

```{r}
data.frame(kraken.cleaning.stram, kraken.cleaning.stram.small) %>%
  select(sort(names(.))) %>%
  mutate(across(everything(), format_large_numbers)) %>% 
  head(7)
```





